Compositional Certified Resource Bounds

### 2015-04-13 paper/main.tex
create Cost Aware Clight, add Q:Tick rule

\and \Rule[leftskip=.5cm, rightskip=.5cm]{Q:Assert}

@@

@@

@@

\caption{Selected inference rules of the quantitative analysis.}

@@

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
$$
%
The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resources are available then the
evaluation does not run out of resources and, if the execution
terminates in state $\state'$, there are at least $Q(\state')$
\mbox{resources left and $\Gamma'$ holds for $\state'$.}

The judgement is a bit more involved since we have to take into
account the early exit statements \code{break} and \code{return}.
This is similar to classical Hoare triples in the presence of non-linear
control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
% of looping statements by the rule {\sc Q:Loop} and used before
% \code{break} statements by the rule {\sc Q:Break}.
Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
returning from a function call with \code{return}. \iffull{(This becomes clear
in the rules.)}{}

@@

We describe the automatic amortized analysis for a subset of
expressions of Clight.  Clight is the first intermediate
language of the CompCert compiler~\cite{CompcertCACM}, it
is a cleaner C with a unified looping construct and
side-effect free expressions.
Assignments are restricted to the form $x \gets
y$ or $x \gets x \pm y$.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This is achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
a[y]$ are handled by assigning $0$ to coefficients
like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
This sound treatment ensures that no further loop bounds
depend on the result of the non-linear operation.

@@

Our current implementation of $\toolname$ does not currently support
all of Clight.  Programs with function pointers, goto statements,
and dereferences of stack-allocated variables cannot be analyzed
automatically.  While these limitations concern the current
implementation, our technique is in principle capable to handle them.

>>>>

\and \Rule{Q:Assert}

@@

\and \Rule{Q:Tick}
{ B, R \vdash \htriple{ \Gamma; Q {+} M_t(n) }{ \code{tick}(n) }{ \Gamma; Q } }

@@

{ (\Gamma_f; Q_f, \Gamma_f'; Q_f') \in \Delta(f) }

@@

@@

\paraskip
\paragraph{Cost Aware Clight.}

We present the rules for a subset of the
expressions of Clight.  Clight is the first intermediate
language of the CompCert compiler~\cite{CompcertCACM}, it
is a cleaner C with a unified looping construct and
side-effect free expressions.  We reuse most of CompCert's
syntax but instrument the semantics with a resource metric
$M$ that accounts for the cost (an arbitrary rational number)
of each step in the operational semantics.  For example, $M_e(\code{exp})$ is the cost of
evaluating the expression \code{exp}; the rationals $M_f$
and $M_r$ account respectively for the cost of a call to the
function $f$ and the cost of returning from it.
More details on this cost-aware semantics are provided in \pref{sec:soundness}.

In the rules, assignments are restricted to the form $x \gets
y$ or $x \gets x \pm y$.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This is achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
a[y]$ are handled by assigning $0$ to coefficients
like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.
This sound treatment ensures that no further loop bounds
depend on the result of the non-linear operation.

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B; Q_B), (\Gamma_R; Q_R) \vdash
    \htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }.
$$
%
The part $\htriple{ \Gamma; Q }{ S }{ \Gamma'; Q' }$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resources are available then the
evaluation does not run out of resources and, if the execution
terminates in state $\state'$, there are at least $Q(\state')$
resources left and $\Gamma'$ holds for $\state'$.

The judgement is a bit more involved since we have to take into
account the early exit statements \code{break} and \code{return}.
This is similar to classical Hoare triples in the presence of non-linear
control flow. In the judgement, $(\Gamma_B; Q_B)$ is the postcondition
that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
% of looping statements by the rule {\sc Q:Loop} and used before
% \code{break} statements by the rule {\sc Q:Break}.
Similarly, $(\Gamma_R; Q_R)$ is the postcondition that holds when
returning from a function call. \iffull{(This becomes clear
in the rules.)}{}

@@

@@

Our current implementation of $\toolname$ does not currently support
all of Clight.  Programs with function pointers, goto statements,
continue statements, and dereferences of stack-allocated variables
cannot be analyzed automatically.  While these limitations concern the
current implementation, our technique is in principle capable to handle
them.
